Effects of Water Permeability on the Hydrodynamics of a Vesicle Under Flow and Confinement

### 2020-12-29 vesicle_code/docs/permeableVes/main.tex

%
%\paragraph{Literature to maybe cite}
%\cite{ali-ovr-ran2020} show how beading can occur on nanotubes. This
%happens in regions where proteins congregate and this can lead to
%semi-permeability?
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htp]
  \centering
  \includegraphics[width=0.45\textwidth]{figures/schematic.pdf}
  \caption{\label{fig:sketch} Schematics showing the effects of
  semipermeability on vesicle hydrodynamics in various configurations.
  The slipper in (a) is the steady-state shape of a semi-permeable
  vesicle in a Poiseuille flow.}
\end{figure}
\section*{Formulation}
We consider a two-dimensional semi-permeable vesicle (permeable only to water) suspended in a viscous fluid, see figure~\ref{fig:sketch}.
To focus on the effect of water flow on vesicle hydrodynamics in various conditions, we assume the fluid is the same inside and outside the vesicle membrane.
%($\gamma$) with an outward normal $\nn$.
On the membrane the permeable water flux gives rise to a difference between the fluid velocity ${\bf u}({\bf x}\in \gamma)$
and the membrane velocity $\dot{\bf x}$ \cite{yao-mor2017}:
\begin{align}
  \label{eqn:vesicleFlux}
  \uu - \dot{\xx} = - k_w (R T \triangle c + \ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\gamma$ denotes the membrane boundary with an outward normal $\nn$, $k_w$ is the water permeability, $R$ is the ideal gas constant, $T$ is the temperature ($^{\circ}$K), $\triangle c$ is contrast in a solute concentration (mol/volume),
and $\ff_\mathrm{mem}$ is the membrane stress that consists of both bending $ \ff_\mathrm{ben}= -k_b \xx_{ssss}$ and tension $\ff_\mathrm{ten}=(\sigma \xx_s)_s$ ($s$
is the arclength along the membrane, see figure~\ref{fig:sketch}. Using $k_w\sim 10^{-12}$ m$^2$s/kg (typical of mammalian cell membranes \cite{LiMoriSun2015_PRL}) and an osmolarity $\triangle c = 0.1$ mol/L (consistent with the osmotic forces in vesicle aspiration experiments \cite{OlbrichRawiczNeedhamEtAl2000_BJ}), the corresponding water flow is of the order $v_\mathrm{o}=0.1 \mu$m/s. Assuming that the membrane bending stiffness $k_b = 10^{-19}$J, a membrane tension $10^{-3}$N/m (one tenth of the lysis tension) and a characteristic length scale of $10^{-6}$m, the tension contribution ($v_\mathrm{t}$) and elastic stress contribution 
($v_\mathrm{e}$) to the permeable water flow can be scaled to $v_\mathrm{o}$ as $v_\mathrm{o}:v_\mathrm{t}:v_\mathrm{e}=1:10^{-2}:10^{-6}$.  Furthermore, we note that in the absence of an osmotic stress, the water permeable flow can be significant when the membrane tension is large due to the external stress such as the hydrodynamics stress due to an external flow and/or due to confinement.
In the following we will explore both.

@@

\section*{Supplementary Material: Formulation \label{sc:formulate}}
We consider a two-dimensional semi-permeable vesicle suspended in an
unbounded viscous fluid driven by the imposed flow $\uu_\infty(\xx)$.
Since fluid is transported across the semi-permeable vesicle membrane,
we only consider fluids with uniform viscosity. Letting $\gamma$ be the
vesicle interface, the fluid velocity $\uu$ and pressure $p$ satisfy
%
\begin{alignat}{3}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \uu &= 0, \quad
  \nabla \cdot \uu = 0, \qquad &&\xx \in \RR^2 \setminus \gamma, \\
%
  \uu(\xx) &\rightarrow \uu_\infty(\xx),
    &&\norm{\xx} \rightarrow \infty,
\end{alignat}
where $\mu$ is the fluid viscosity. Letting $\nn$ be the outward normal
to $\gamma$, $\jump{\cdot}$ be the jump across the interface, and $T$ be
the hydrodynamic stress tensor, the interfacial boundary conditions are
\begin{align}
  \label{eqn:jump}
  \jump{T\cdot\nn} = \ff_\mathrm{mem}, \quad \jump{\uu} = 0,
\end{align}
where $\ff_\mathrm{mem}$ is the membrane force that results from the
Helfrich energy with bending modulus $k_b$, and a tension term that acts
as a Lagrange multiplier to satisfy the local inextensibility condition.
Following~\cite{vee-gue-zor-bir2009}, we modify the bending tension and
forces as
$\ff_\mathrm{mem} = \ff_\mathrm{ben} +
\ff_\mathrm{ten}$, where
\begin{align}
  \label{eqn:membraneForces}
  \ff_\mathrm{ben} = -k_b \xx_{ssss}, \qquad
  \ff_\mathrm{ten} = (\sigma \xx_s)_s.
\end{align}
The physical tension is $\Lambda = \sigma + 1.5 \kappa^2$, where
$\kappa$ is the membrane curvature.

Semi-permeability is introduced by modifying the vesicle velocity from
the standard choice. Instead of using the no-slip boundary condition, we
assume the fluid flux through the boundary is proportional to the
mechanical force~\cite{yao-mor2017}. That is, 
\begin{align}
  \label{eqn:vesicleFlux}
  \uu - \dot{\xx} = - \beta (\ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\beta$ is the semi-permeability rate.

>>>>

%
%\paragraph{Literature to maybe cite}
%\cite{ali-ovr-ran2020} show how beading can occur on nanotubes. This
%happens in regions where proteins congregate and this can lead to
%semi-permeability?
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htp]
  \centering
  \includegraphics[width=0.45\textwidth]{figures/schematic.pdf}
  \caption{\label{fig:sketch} Schematics showing the effects of
  semipermeability on vesicle hydrodynamics in various configurations.
  The slipper in (a) is the steady-state shape of a semi-permeable
  vesicle in a Poiseuille flow.}
\end{figure}

\section*{Formulation}
We consider a two-dimensional semi-permeable vesicle (permeable only to water) suspended in a viscous fluid, see figure~\ref{fig:sketch}.
To focus on the effect of water flow on vesicle hydrodynamics in various conditions, we assume the fluid is the same inside and outside the vesicle membrane.
%($\gamma$) with an outward normal $\nn$.
On the membrane the permeable water flux gives rise to a difference between the fluid velocity ${\bf u}({\bf x}\in \gamma)$
and the membrane velocity $\dot{\bf x}$ \cite{yao-mor2017}:
\begin{align}
  \label{eqn:vesicleFlux}
  \uu - \dot{\xx} = - k_w (R T \triangle c + \ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\gamma$ denotes the membrane boundary with an outward normal $\nn$, $k_w$ is the water permeability, $R$ is the ideal gas constant, $T$ is the temperature ($^{\circ}$K), $\triangle c$ is contrast in a solute concentration (mol/volume),
and $\ff_\mathrm{mem}$ is the membrane stress that consists of both bending $ \ff_\mathrm{ben}= -k_b \xx_{ssss}$ and tension $\ff_\mathrm{ten}=(\sigma \xx_s)_s$ ($s$
is the arclength along the membrane, see figure~\ref{fig:sketch}. Using $k_w\sim 10^{-12}$ m$^2$s/kg (typical of mammalian cell membranes \cite{LiMoriSun2015_PRL}) and an osmolarity $\triangle c = 0.1$ mol/L (consistent with the osmotic forces in vesicle aspiration experiments \cite{OlbrichRawiczNeedhamEtAl2000_BJ}), the corresponding water flow is of the order $v_\mathrm{o}=0.1 \mu$m/s. Assuming that the membrane bending stiffness $k_b = 10^{-19}$J, a membrane tension $10^{-3}$N/m (one tenth of the lysis tension) and a characteristic length scale of $10^{-6}$m, the tension contribution ($v_\mathrm{t}$) and elastic stress contribution 
($v_\mathrm{e}$) to the permeable water flow can be scaled to $v_\mathrm{o}$ as $v_\mathrm{o}:v_\mathrm{t}:v_\mathrm{e}=1:10^{-2}:10^{-6}$.  Furthermore, we note that in the absence of an osmotic stress, the water permeable flow can be significant when the membrane tension is large due to the external stress such as the hydrodynamics stress due to an external flow and/or due to confinement.
In the following we will explore both.

@@

\section*{Supplementary Material}
\subsection*{Formulation}
We consider a two-dimensional inextensible semi-permeable vesicle,
$\gamma$, suspended in unconfined and confined geometries. The
suspending fluid is assumed to be viscous so that the fluid velocity
$\uu$ and pressure $p$ are governed by the incompressible Stokes
equations
\begin{align}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \uu = 0, \quad
  \nabla \cdot \uu = 0, \qquad \xx \in \RR^2 \setminus \gamma,
\end{align}
where $\mu$ is the fluid viscosity. In an unconfined geometry, the
condition $\uu(\xx) \rightarrow \uu_\infty(\xx)$ is enforced as $\xx
\rightarrow \infty$, and in a confined geometry with boundary $\Gamma$,
the boundary condition $\uu(\xx) = \UU(\xx)$ is enforced on $\Gamma$,
where $\UU(\xx)$ is prescribed. Along the vesicle $\gamma$, mass
continuity, a force balance, and local inextensibility are enforced by
requiring
\begin{align}
  \label{eqn:jump}
  \jump{\uu} = 0, \quad
  \jump{T\cdot\nn} = \ff_\mathrm{mem}, \quad
  \xx_s \cdot \uu_s = 0,
\end{align}
where $\nn$ is the outward normal of $\gamma$, $\jump{\cdot}$ is the
jump across the interface, and $T$ is the hydrodynamic stress tensor.
The membrane force $\ff_\mathrm{mem}$ is the sum of a bending force
with bending modulus $k_b$ and a force due to the tension $\Lambda$.
Following~\cite{vee-gue-zor-bir2009}, we modify the bending and tension
forces as $\ff_\mathrm{ben} = -k_b \xx_{ssss}$, and $\ff_\mathrm{ten} =
(\sigma \xx_s)_s$, where $s$ is arclength. The physical tension can be
recovered using the identity $\Lambda = \sigma + 1.5 \kappa^2$, where
$\kappa$ is the membrane curvature.

Semi-permeability is introduced by modifying the vesicle velocity from
the standard choice. Instead of using the no-slip boundary condition, we
assume the fluid flux through the boundary is proportional to the
mechanical force~\cite{yao-mor2017}. That is, 
\begin{align}
  \label{eqn:vesicleFlux}
  \uu - \dot{\xx} = - k_w (\ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $k_w$ is the water permeability.

The governing equations are non-dimensionalized by introducing a length
scale that we take to be $R_0 = 10^{-6}$m. Using a bending stiffness of
$k_b = 10^{-19}$J and fluid viscosity $5 \times 10^{-2}$kg/ms, we define
the time scale $\tau = \mu R_0^3/k_b = 0.5s$, the velocity scale $U =
R_0/\tau = 2\mu$m/s, the pressure scale $P = k_b/R_0^3 = 10^{-1}$Pa, the
tension scale $Q = k_b/R_0 = 10^{-13}$N, and the permeability scale $W =
R_0\mu = 2 \times 10^{-5}$m$^2$s/kg. The other standard dimensionless
variable of a vesicle is its reduced area $\nu = 4\pi A/L^2 \in (0,1]$.
The length $L$ is conserved by the local inextensibility condition.
However, the area $A$ is not conserved because of semi-permeability, and
therefore $\nu$ is also not conserved.
